---
title: "Data Exploration Plots"
author: "Rui Nascimento"
date: "`r format(Sys.time(), '%d %B, %Y (%X)')`"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r loadLibs}
library(data.table)
library(ggplot2)
library(magrittr)
```

# Raw Data Plots for data exploration

Load Raw counts_data from .RDS file and melt the dataset

```{r}
counts_data <- readRDS("data/counts_data.RDS")

log_counts_data <- log2(counts_data)

df <- data.table::melt(data = log_counts_data,
           value.name = "Counts",
           variable.name = "Run",
           measure.vars = names(counts_data))

metadata <- data.table(readRDS("data/pinot.RDS"), keep.rownames = TRUE)
setnames(metadata, "rn", "Run")

final <- data.table(merge.data.table(x = df, y = metadata, by = "Run"))
```

## By LibLayout

Raw counts colored by LibLayout

```{r}
final %>%
ggplot(aes(x = Run, y = Counts, fill = LibraryLayout)) + geom_boxplot() + xlab("") +
  ylab(expression(log[2](raw_counts))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Raw Counts by Lybrary Layout")
```

## By Stress

Raw counts colored by Stress

```{r}
final %>%
  ggplot(aes(x = Run, y = Counts, fill = Stress)) + geom_boxplot() + xlab("") +
  ylab(expression(log[2](raw_counts))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Raw Counts by Stress")
```

## By BioProject

Raw counts colored by BioProject

```{r}
final %>%
  ggplot(aes(x = Run, y = Counts, fill = BioProject)) + geom_boxplot() + xlab("") +
  ylab(expression(log[2](raw_counts))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Raw Counts by Bioproject")
```

## Density

```{r}
final %>%
  ggplot(aes(x = Counts, colour = Run, fill = Run)) +
  geom_density(alpha = 0.1) +
  theme(legend.position = "none") + xlab(expression(log[2](count + 1))) +
  labs(title="Density Plot of Raw Counts")
```

# Normalized counts with limma-voom

Read normalized counts

```{r}
logcpm <- readRDS(file = "data/logcpm.RDS")
```

Melt the dataset

```{r}
df_cpm <- data.table::melt(data = logcpm,
                       value.name = "Counts",
                       variable.name = "Run",
                       measure.vars = names(logcpm))
setnames(df_cpm, c("Var1", "Var2"), c("ID", "Run"))
final_norm <- data.table(merge.data.table(x = df_cpm, y = metadata, by = "Run"))
```

## By LibLayout

Normalized counts colored by LibLayout

```{r}
final_norm %>%
ggplot(aes(x = Run, y = Counts, fill = LibraryLayout)) + geom_boxplot() + xlab("") +
  ylab(expression(normalied_counts)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Normalized Counts by Lybrary Layout")
```

## By Stress

Normalized counts colored by Stress

```{r}
final_norm %>%
  ggplot(aes(x = Run, y = Counts, fill = Stress)) + geom_boxplot() + xlab("") +
  ylab(expression(normalied_counts)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Normalized Counts by Stress")
```

## By BioProject

Normalized counts colored by Bioproject

```{r}
final_norm %>%
  ggplot(aes(x = Run, y = Counts, fill = BioProject)) + geom_boxplot() + xlab("") +
  ylab(expression(normalied_counts)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Normalized Counts by Bioproject")
```

## Density

```{r}
final_norm %>%
  ggplot(aes(x = Counts, colour = BioProject, fill = Run)) +
  geom_density(alpha = 0.1) +
  theme(legend.position = "none") + xlab(expression(normalied_counts)) +
  labs(title="Density Plot of Normalized Counts")
```

## Plot a Random number of Runs

```{r}
number_of_samples = 10

final_norm[Run %in% sample(unique(final_norm$Run), number_of_samples)] %>%
  ggplot(aes(x = Counts, colour = Run, fill = BioProject)) +
  geom_density(alpha = 0.1) +
  theme(legend.position = "right") + xlab(expression(normalied_counts)) +
  labs(title="Density Plot of Normalized Counts")
```
